Evaluating the Effects of Tracker Reliability and Field of View on a Target Following Task in Augmented Reality

### 2009-06-25 Paper/paper.tex
Added independent variable notation

We examine the effects of level of immersion and tracking sensor reliability in augmented reality (AR) X-ray vision with a virtual reality (VR) based simulation.  We analyze participant performance on a target following task while varying the field of view of the AR display, as well as the reliability of the head tracking sensor.  In low reliability conditions, we simulate sensor dropouts by disabling the augmented view of the scene for brief time periods.  Our study gives insight into the effect of tracking sensor reliability on user performance in a target following task, as well as the relationship between reliability and field of view in an AR system.

@@

In our experiment we vary two parameters: the field of view of the AR interface and the length of sensor dropout periods.  Each trial is 60 seconds, with seven sensor dropouts.  The total vertical field of view of the HMD is 36 degrees; the three possible values for the augmented field of view were 10, 20, and 34 degrees.  The length of dropouts varied between two seconds (highest), one second (medium) and zero seconds (lowest).  We experimented with longer dropout periods but deemed them too long.  This gave us a total of 9 conditions, and for each participant we tested each condition 3 times giving a total of 27 trials per participant.

@@

From this plot we can see that with a small field of view, performance is equally bad with no dropouts or one second dropouts.  The small field of view seems to have a dominating effect on performance.  When the field of view is increased to the medium level, performance dramatically increases when there are no dropouts, but only gets slightly better when there are dropouts.  Finally, we do not observe much difference in performance between the medium and large field of view cases.  Using Tukey's HSD to compare means, the medium and large field of view conditions are in fact not significantly different (p=??).

\begin{figure}[t]
	\centering
	\includegraphics[width=3.5in]{figures/numtimes_interaction.pdf}
	\caption{\label{fig:interaction}Interaction plot of the effect of field of view and sensor dropout length on the number of times the participant loses the target.}
\end{figure}

@@

Figure \ref{fig:tt_deadlen} shows a box plot of time following target versus dropout length.  Performance decreases between zero and one seconds, but seems to level off between one and two.  We ran a post-hoc analysis to determine which conditions were significantly different.  Using Tukey's HSD, we found that conditions one and two were significantly different than zero (p=??), but not significantly different from each other (p=??).  This suggests that we may be seeing a thresholding effect above one second sensor dropouts.
% TODO: table with values

To investigate this threshold, we had seven participants complete an extra three trials with the medium field of view and sensor dropout periods of 0.5 seconds.  Figure \ref{fig:extra} shows a box plot of time following target versus sensor dropout length for these seven participants.  Again using Tukey's HSD to compare the means, we did not find a significant difference between the zero length and half second length conditions, but did find a significant difference between those and the higher length conditions.  
%TODO: table with values
With further experimentation, we could further delineate the performance curve as the dropout length increases from zero to one.
%For future work, we could find the threshold between 0.5 and 1.

\begin{figure}[htb]
	\centering
	\includegraphics[width=3.5in]{figures/extra.pdf}
	\caption{Box plot of the number of times the participant loses the target versus the length of the sensor dropout periods, including the half second dropout condition.  We collected this data from seven participants, using the medium field of view.}
\end{figure}

@@

>>>>

We examine the effects of level of immersion and tracking sensor reliability in augmented reality (AR) X-ray vision with a virtual reality (VR) based simulation.  We analyze participant performance on a target following task while varying the field of view of the AR display, as well as the reliability of the head tracking sensor.  In low reliability conditions, we simulate sensor dropouts by disabling the augmented view of the scene for brief time periods.  Our study gives insight into the effect of tracking sensor reliability on user performance in a target following task, as well as the relationship between sensor reliability and field of view in an AR system.

@@

In our experiment we have two independent variables: the field of view of the AR interface and the length of sensor dropout periods.  Each trial is 60 seconds, with seven sensor dropouts.  The total vertical field of view of the HMD is 36 degrees; the three possible values for the augmented field of view were 10, 20, and 34 degrees.  The length of dropouts varied between two seconds (highest), one second (medium) and zero seconds (lowest).  We experimented with longer dropout periods but deemed them too long.  This gave us a total of 9 conditions, and for each participant we tested each condition 3 times giving a total of 27 trials per participant.

@@

From this plot we can see that with a small field of view, performance is equally bad with no dropouts or one second dropouts.  The small field of view seems to have a dominating effect on performance.  When the field of view is increased to the medium level, performance dramatically increases when there are no dropouts, but only gets slightly better when there are dropouts.  Finally, we do not observe much difference in performance between the medium and large field of view cases.  Using Tukey's HSD to compare means, the medium and large field of view conditions are in fact not significantly different ($p=0.575$).

\begin{figure}[t]
	\centering
	\includegraphics[width=2.5in]{figures/numtimes_interaction.pdf}
	\caption{\label{fig:interaction}Interaction plot of the effect of field of view and sensor dropout length on the number of times the participant loses the target.}
\end{figure}

@@

Figure \ref{fig:tt_deadlen} shows a box plot of time following target versus dropout length.  Performance decreases between zero and one seconds, but seems to level off between one and two.  We ran a post-hoc analysis to determine which conditions were significantly different.  Using Tukey's HSD, we found that the one second and two second conditions were significantly different than zero ($p<0.001$ for both), but not significantly different from each other ($p=0.239$).  This suggests that we may be seeing a thresholding effect above one second sensor dropouts.
% TODO: table with values

To investigate this threshold, we had seven participants complete an extra three trials with the medium field of view and sensor dropout periods of 0.5 seconds.  Figure \ref{fig:extra} shows a box plot of time following target versus sensor dropout length for these seven participants.  Again using Tukey's HSD to compare the means, we did not find a significant difference between the zero length and half second length conditions, but did find a significant difference between those and the higher length conditions.
%TODO: table with values
With further experimentation, we could further delineate the performance curve as the dropout length increases from zero to one.

\begin{figure}[htb]
	\centering
	\includegraphics[width=2.5in]{figures/extra.pdf}
	\caption{\label{fig:extra}Box plot of the total time the participant is following the target versus the length of the sensor dropout periods, including the half second dropout condition.  We collected this data from seven participants, using the medium field of view.}
\end{figure}

@@

We designed an experiment to study a target following task which makes use of X-ray vision through an augmented reality device.  We used an augmented reality simulation to make the experiment feasible.

We found that for our task, a low field of view has an adverse effect on performance.  
%This adverse effect dominates the other variable component of immersion, the reliability of the head-based rendering.
With reliable head-based rendering, a medium field of view is enough for good performance.  We did not see a significant performance increase when increasing to a high field of view.

Decreased reliability in the head-based rendering, caused by periods of sensor dropout, adversely affects performance.  However, for our task, the performance decrease levels off at one second dropouts.  Preliminary analysis also shows no significant difference between having no dropouts and half second dropouts.  This suggests that below a certain length of dropouts, good level of performance can be maintained, provided that the field of view is large enough.  This threshold on the dropouts would change depending on what is being followed, but we expect the same pattern would appear in other target following scenarios.

Future work lies in further investigating the effect of the dropout length 
below the one second level.  Also, this work points to other interesting experiments which might test the interaction between reliability of the head-based rendering and other immersion components.
